Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2020-01-12 rnni_convexity.tex
typo
add Mike's suggestions

\newtheorem{definition}{Definition}
\newtheorem{theorem}[definition]{Theorem}
\newtheorem{conjecture}[definition]{Conjecture}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{proposition}[definition]{Proposition}

@@

\title[Ranked Nearest Neighbour Intarchange]{Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees}
\date{\today}
\author{Lena Collienne\textsuperscript{1}}
\email{lena.collienne@postgrad.otago.ac.nz}
\address{\textsuperscript{1}Department of Computer Science, University of Otago, New Zealand}
\author{Kieran Elmes\textsuperscript{1}}
\email{kelmes@cs.otago.ac.nz}
\author{Mareike Fischer\textsuperscript{2}}
\email{email@mareikefischer.de}
\address{\textsuperscript{2}Institute of Mathematics and Computer Science, University of Greifswald, Germany}
\author{David Bryant\textsuperscript{3}}
\email{david.bryant@otago.ac.nz}
\address{\textsuperscript{3}Department of Mathematics and Statistics, University of Otago, New Zealand}
\author{Alex Gavryushkin\textsuperscript{1, \Letter}}
\email{\textsuperscript{\Letter}alex@biods.org}

@@

Case $k > 1$.
\begin{enumerate}
\item $C_k = C_i \cup C_j$ for $i, j < k$.
Suppress $C_i$ and $C_j$ in both $T'$ and $R$ to new taxa $c_i$ and $c_j$ respectively and proceed as in Case $k = 1$.
\item $C_k = C_i \cup {x}$, where $x$ is a taxon not present in $C_1, \ldots, C_k$.
Suppress $C_i$ in both $T'$ and $R$ to a new taxon $c_i$ and proceed as in Case $k = 1$.
\item $C_k = \{x, y\}$ where both $x$ and $y$ are not in $C_1, \ldots, C_k$.
Proceed as in Case $k = 1$.
\end{enumerate}
\endproof

@@

\begin{lemma}
Let $T$ and $R$ be two trees on taxa $\{1, \ldots, n+1\}$.
Then $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta$, where $\delta = |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
\label{lemma:distance_delete_taxon}
\end{lemma}

@@

\textcite[Theorem~7]{Gavryushkin2018-ol} gave an upper bound for the diameter of $\rnni$ space, denoted by $\Delta(\rnni)$.
They showed that $\Delta(\rnni) = n^2 - 3n - 5/8$.
In this section we improve that result and calculate the exact diameter and radius of the $\rnni$ space.

@@

\begin{conjecture}[Weak Split Theorem]
If a split of taxa $A | B$ is shared by two trees $T$ and $R$ in $\rnni$ then there exists a shortest path $p$ between $T$ and $R$ such that all trees on $p$ share the split $A | B$.
\label{conjecture:split_theorem}
\end{conjecture}

@@

\begin{conjecture}[Cluster Theorem]
Let $T$ and $R$ be trees in $\rnni$ that share a cluster $C$.
Then $C$ is present as a cluster in every tree on every shortest path between $T$ and $R$.
\label{conjecture:cluster_theorem}
\end{conjecture}

>>>>

\newtheorem*{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}

@@

\title[Ranked Nearest Neighbour Interchange]{Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees}
\date{\today}
\author{Lena Collienne\textsuperscript{1}}
\email{lena.collienne@postgrad.otago.ac.nz}
\address{\textsuperscript{1}Department of Computer Science, University of Otago, New Zealand}
\author{Kieran Elmes\textsuperscript{1}}
\email{kelmes@cs.otago.ac.nz}
\author{Mareike Fischer\textsuperscript{2}}
\email{email@mareikefischer.de}
\address{\textsuperscript{2}Institute of Mathematics and Computer Science, University of Greifswald, Germany}
\author{David Bryant\textsuperscript{3}}
\email{david.bryant@otago.ac.nz}
\address{\textsuperscript{3}Department of Mathematics and Statistics, University of Otago, New Zealand}
\author{Alex Gavryushkin\textsuperscript{1, \Letter}}
\email{\textsuperscript{\Letter}alex@biods.org}

@@

Case $k > 1$.
\begin{enumerate}
\item $C_k = C_i \cup C_j$ for $i, j < k$.
Suppress $C_i$ and $C_j$ in both $T'$ and $R$ to new taxa $c_i$ and $c_j$ respectively and proceed as in Case $k = 1$.
\item $C_k = C_i \cup \{x\}$, where $x$ is a taxon not present in $C_1, \ldots, C_k$.
Suppress $C_i$ in both $T'$ and $R$ to a new taxon $c_i$ and proceed as in Case $k = 1$.
\item $C_k = \{x, y\}$ where both $x$ and $y$ are not in $C_1, \ldots, C_k$.
Proceed as in Case $k = 1$.
\end{enumerate}
\endproof

@@

\begin{lemma}
Let $T$ and $R$ be two trees on taxa $\{1, \ldots, n+1\}$.
Then ${d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - \delta}$, where $\delta = |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))|$.
\label{lemma:distance_delete_taxon}
\end{lemma}

@@

\textcite[Theorem~7]{Gavryushkin2018-ol} gave an upper bound for the diameter of $\rnni$ space, denoted by $\Delta(\rnni)$.
They showed that $\Delta(\rnni) \leq n^2 - 3n - 5/8$.
In this section we improve that result and calculate the exact diameter and radius of the $\rnni$ space.

@@

\begin{definition}[Weak Split Theorem]
If a split of taxa $A | B$ is shared by two trees $T$ and $R$ then there exists a shortest path $p$ between $T$ and $R$ such that all trees on $p$ share the split $A | B$.
\end{definition}

\begin{conjecture}
\label{conjecture:weak_split_theorem}
The Weak Split Theorem holds for $\rnni$.
\end{conjecture}

@@

\begin{definition}[Cluster Theorem]
Let $T$ and $R$ be trees that share a cluster $C$.
Then $C$ is present as a cluster in every tree on every shortest path between $T$ and $R$.
\end{definition}

\begin{conjecture}
\label{conjecture:cluster_theorem}
The Cluster Theorem holds for $\rnni$.
\end{conjecture}
